Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2012-09-06 Dail_Mad_ext.tex
major revisions of intro

A third problem with these models is that they do not admit
spatial variation, which
%. Recent work has sought to remedy this deficiencyThis is not a fair criticism because this
%lies beyond the scope of traditional state-space models, but
%it does
restrict the utility of their since it is typically
impossible assess the impacts of factors such as habitat
fragmentation or climate change on population dynamics without
considering spatial variation in abundance. Indeed many populations are
regulated by spatial processes such as source-sink dynamics.

Although numerous extensions have been proposed to allow for spatial
variation in abundance, in this paper we focus on the model of
\citet{dail_madsen:2011} developed a model (henceforth the DM model)
that simultaneous resolves each of these problems with traditional state-space
models. Specifically, the model (1) recognizes the discrete nature of
count data and the underlying true abundance state, (2) has estimable
parameters, including those describing the initial state at time
$t=1$, and (3) allows for inference about both temporal and spatial
variation in abundance.
The DM model allows for inference about spatial and temporal
dynamics in abundance while accounting for observation error using
only spatially and temporally replicated count data.  Their model is
an extension of the closed-population $N$-mixture model
\citep{royle:2004biom}, which was designed specifically to address the
problem of
modeling spatial variation in abundance in the presence of observation
error, specifically, when detection probability is less than one. The
DM extension relaxes the assumption of population
closure and includes explicit parameters describing population change
as a first-order Markovian process. [more details]

@@

\section{Model Development}

\subsection{Original Formulation}

@@

@@

\section{Applications}

>>>>

A fourth problem with these models is that they do not admit
spatial variation, which
%. Recent work has sought to remedy this deficiencyThis is not a fair criticism because this
%lies beyond the scope of traditional state-space models, but
%it does
restrict the utility of their since it is typically
impossible assess the impacts of factors such as habitat
fragmentation or climate change on population dynamics without
considering spatial variation in abundance. Indeed many populations are
regulated by spatial processes such as source-sink dynamics.

Several extensions of state-space models have been proposed to
overcome the limitations described
above. \citet{devalpine_hastings:2002} described methods for fitting
models with non-Gaussian distributions for the process and observation
errors. Observation models with more intuitive interpretations have
been proposed by \citet{kery_etal:2009}. \citet{lele_etal:1998} developed
models allowing for inference about spatial and temporal variation in
abundance, and their developments also resolved the problems of
non-identifiability for the parameters of the initial state at time
$t=1$. In spite of these improvements, until recently, few methods
existed to address each issue simultaneously, and even fewer of these
methods have been made readily available to applied ecologists.

In this paper, we focus on the model of
\citet{dail_madsen:2011} %developed a model (henceforth the DM model)
that simultaneous resolves each of the abovementioned problems with
traditional state-space models. Specifically, the model (1) recognizes
the discrete nature of count data and the underlying true abundance
state, (2) includes a realistic observation model with explicit
detection probability parameters (3) has estimable parameters,
including those describing the initial state at time $t=1$, and (4)
allows for inference about both temporal and spatial variation in
abundance. In some respects, the DM model can be viewed as a specific
manifestation of the general state-space model formulated by
\citep{buckland_etal:2004}. However, their model was designed to allow
for inference about demographic parameters, which will typically
require mark-recapture data. Although this may be viewed as the
gold standard in many studies of wildlife populations, in this paper,
we focus on the analysis of count data, which is the typical source of
information used in state-state modeling applications, and one of the
most widely available forms of ecological data---it is also the data
required by the DM model.

In the following section, we describe the DM model in its original
form, and we explain how it resolves each of the deficieinces with
standard state-space models. In Section~\ref{sec:ext}, we extend the
model to accomodate classical models of population growth and to
handle several features common to ecological
time-series. Specifically, we describe methods for accomodating excess
zeros and nuisance variables such as random observer effects. Both
frquentist and Bayesian methods of inference are discussed, and code
for fitting models is presented in the appendices. The Bayesian method
is attractive in that it can accomodate prior information about
detection parameters, which is available for many species, yet not
collected as part of many existing monitoring programs. In
Section~\ref{sec:app}, we evaluate the performance of the model
extensions using a simulation study and by analyzing data from the
North American Breeding Bird Survey (BBS), one of
the most spatially and temporally extensive sets of count data on
vertebrate populations \citep{robbins_etal:1986}. The overarching aim
of  the paper is to provide ecologists with a flexible and accessible means of
adressing important questions related to the variaion of abundance in
space and time.
believe these extensions will increase the utility of the model for
application to existing datasets such the Breeding Bird Survey.

\begin{comment}
  The DM model allows for inference about spatial and temporal
  dynamics in abundance while accounting for observation error using
  only spatially and temporally replicated count data.  Their model is
  an extension of the closed-population $N$-mixture model
  \citep{royle:2004biom}, which was designed specifically to address
  the problem of modeling spatial variation in abundance in the
  presence of observation error, specifically, when detection
  probability is less than one. The DM extension relaxes the
  assumption of population closure and includes explicit parameters
  describing population change as a first-order Markovian
  process. [more details]

@@

  In this paper, we demonstrate how the DM model resolves each of
  these four limitations of conventional state-space models, and we
  extend the DM model in several important ways. First, we demonstrate
  how classical population models can be embedded within the DM
  model. Then we present methods for handling excess zeros as well as
  nuisance parameters such as random observer effects on detection
  probability. A simulation study
\end{comment}

\section{The Dail-Madsen Model}
\label{sec:dm}

%\subsection{Original Formulation}

@@

\section{Model Extensions}
\label{sec:ext}

@@

\section{Applications}
\label{sec:app}
